---
output: 
    md_document:
        variant: gfm
---

```{r, echo = FALSE}
knitr::opts_chunk$set(
  fig.path = "README_figs/README-"
)
```

# ubms: Unmarked Bayesian Models with Stan

[![Build Status](https://travis-ci.org/kenkellner/ubms.svg?branch=master)](https://travis-ci.org/kenkellner/ubms)
[![codecov](https://codecov.io/gh/kenkellner/ubms/branch/master/graph/badge.svg)](https://codecov.io/gh/kenkellner/ubms)

In-development `R` package for fitting Bayesian hierarchical models of animal occurrence and abundance.
The package has a formula-based interface compatible with [unmarked](https://cran.r-project.org/web/packages/unmarked/index.html), but the model is fit using MCMC with [Stan](https://mc-stan.org/) instead of using maximum likelihood.
Right now there are Stan versions of unmarked functions `occu`, `occuRN`, `colext`, and `pcount` (`stan_occu`, `stan_occuRN`, `stan_colext`, `stan_pcount`).

Advantages compared to `unmarked`:

  1. Obtain posterior distributions of parameters and derived parameters
  2. Include random effects in parameter formulas (same syntax as `lme4`)

Disadvantages compared to `unmarked`:

  1. MCMC is slower than maximum likelihood
  2. Limited selection of model types
  3. Potential convergence issues

### Example

Simulate occupancy data including a random effect on occupancy:

```{r,message=F,warning=F}
library(ubms)

set.seed(123)
dat_occ <- data.frame(x1=rnorm(500))
dat_p <- data.frame(x2=rnorm(500*5))

y <- matrix(NA, 500, 5)
z <- rep(NA, 500)

b <- c(0.4, -0.5, 0.3, 0.5)

re_fac <- factor(sample(letters[1:26], 500, replace=T))
dat_occ$group <- re_fac
re <- rnorm(26, 0, 1.2)
re_idx <- as.numeric(re_fac)

idx <- 1
for (i in 1:500){
  z[i] <- rbinom(1,1, plogis(b[1] + b[2]*dat_occ$x1[i] + re[re_idx[i]]))
  for (j in 1:5){
    y[i,j] <- z[i]*rbinom(1,1, 
                    plogis(b[3] + b[4]*dat_p$x2[idx]))
    idx <- idx + 1
  }
}
```

Create `unmarked` frame:

```{r}
umf <- unmarkedFrameOccu(y=y, siteCovs=dat_occ, obsCovs=dat_p)
```

Fit a model with a random intercept:

```{r}
options(mc.cores=3) #number of parallel cores to use
(fm <- stan_occu(~x2 ~x1 + (1|group), umf, refresh=0))
```

Examine residuals for occupancy and detection submodels (following [Wright et al. 2019](https://doi.org/10.1002/ecy.2703)).
Each panel represents a draw from the posterior distribution.

```{r resids, fig.height=7}
plot(fm)
```

Assess model goodness-of-fit with a posterior predictive check, using the MacKenzie-Bailey chi-square test:

```{r gof}
(fm_fit <- gof(fm, draws=500, quiet=TRUE))
plot(fm_fit)
```
